### 2016-04-25 Paper_v5.tex
Updated Paper

Meta-heuristic stochastic optimization approaches like Genetic Algorithms (GAs) \cite{goldberg2006genetic}, Simulated Annealing (SA) \cite{kirkpatrick1983optimization},
Evolutionary Programming \cite{fogel2009artificial} and population based searches like Differential Evolution (DE) \cite{storn1997differential} have shown promise on nonlinear multi-modal problems \cite{sun2012parameter}.
These techniques do not make any assumptions about the structure of objective function nor do they require \textit{a~priori} information about the objective function.
Though they do not guarantee strong convergence, these approaches are effective in finding near optimal solutions.
Mendes et al. \cite{mendes1998non} used simulated annealing to estimate rate constants for the irreversible inhibition of HIV proteinase,
Modchang et al. \cite{modchang2008mathematical} used a genetic algorithm (GA) to estimate parameters for a model of G-protein-coupled receptors(GPCRs) mediated signal transduction. The parameters obtained using GA helped qualitatively ascertain the efficacy and potency of two G-protein agonists - CPA and NECA. 
Differential evolution approaches have also been effective on various biological problems \cite{tsai2005evolutionary,wang2001hybrid,noman2007inferring}.
Tashkova et al. compared different meta-heuristics for parameter estimation on a dynamic model of endocytosis and showed that DE was the most effective \cite{tashkova2011parameter}.
Banga and co-workers have also successfully applied scatter-search methods to estimate parameters on non-linear biological processes \cite{villaverde2012cooperative,rodriguez2006novel,egea2007scatter}.
Hybrid approaches that combine a meta-heuristic with a local optimization search, wherein a near globally optimal solution obtained using a meta-heuristic is refined using a local search have also become popular.
Villaverde et al. combined scatter search with local search methods for parameter estimation in a collection of systems biology models \cite{villaverde2015biopredyn}.
Fan et al. recently showed that population based meta-heuristics along with decomposition based methods can be also used to model gene circuits from mRNA data \cite{fan2015parameter}.
Despite these successes, a major drawback of most metaheuristic approaches is the large number of function evaluations required to explore the parameter space.
Typically as models grow in size and complexity, evaluation of the objective function becomes computationally expensive.
Thus performing a large number of function evaluations is not desirable (and perhaps not feasible).

@@

Villaverde and co-workers published a set of benchmark biochemical problems to evaluate parameter estimation methods \cite{villaverde2015biopredyn}.
They ranked ordered these example problems by computational cost from most to least expensive.
We evaluated the performance of DOPS on problems from the least and most expensive categories.
The least expensive problem (henceforth referred to as problem B4) was a metabolic model of Chinese Hamster Ovary (CHO) with 35 metabolites, 32 reactions and 117 parameters.
The biochemical reactions were modeled using modular rate laws and the generalized form of Michaelis-Menten kinetics.
The expensive problem was a genome scale kinetic model of \textit{Saccharomyces~cerevisiae} with 261 reactions, 262 variables and 1759 parameters (henceforth referred to as problem B1). In both cases, synthetic time series data sets were generated using known parameter values, these data were then used by the parameter estimation methods.
For problem B1, the time series data consisted of 44 observables, and for problem B4 the data corresponded to 13 different metabolite measurement sets.
We fixed the number of function evaluations at $\mathcal{N}$ = 4000, and trained both models against the synthetic experimental data.
In both cases we found good fits  (Fig. \ref{fig-sims-b1} and Fig. \ref{fig-sims-b4}) to the problems.
We also recaptured the nominal parameter values within the specified number of function evaluations (Fig. \ref{fig-benchmark}).
Thus, DOPS estimated the parameters in benchmark biochemical models, and recovered the original parameters from synthetic data.
Next, we compared the performance of DOPS with the four other meta-heuristic approaches for a model of the human coagulation cascade.

>>>>

Meta-heuristic stochastic optimization approaches like Genetic Algorithms (GAs) \cite{goldberg2006genetic}, Simulated Annealing (SA) \cite{kirkpatrick1983optimization},
Evolutionary Programming \cite{fogel2009artificial} and population based searches like Differential Evolution (DE) \cite{storn1997differential} have shown promise on nonlinear multi-modal problems \cite{sun2012parameter}.
These techniques do not make any assumptions about the structure of objective function nor do they require \textit{a~priori} information about the objective function.
Though they do not guarantee strong convergence, these approaches are effective in finding near optimal solutions.
Mendes et al. \cite{mendes1998non} used simulated annealing to estimate rate constants for the irreversible inhibition of HIV proteinase,
Modchang et al. \cite{modchang2008mathematical} used a genetic algorithm (GA) to estimate parameters for a model of G-protein-coupled receptors (GPCRs) mediated signal transduction. The parameters obtained using GA helped qualitatively ascertain the efficacy and potency of two G-protein agonists - N6-cyclopentyladenosine (CPA) and 5'-N-ethylcarboxamidoadenosine (NECA). 
Differential evolution approaches have also been effective on various biological problems \cite{tsai2005evolutionary,wang2001hybrid,noman2007inferring}.
Tashkova et al. compared different meta-heuristics for parameter estimation on a dynamic model of endocytosis and showed that DE was the most effective \cite{tashkova2011parameter}.
Banga and co-workers have also successfully applied scatter-search methods to estimate parameters on non-linear biological processes \cite{villaverde2012cooperative,rodriguez2006novel,egea2007scatter}.
Hybrid approaches that combine a meta-heuristic with a local optimization search, wherein a near globally optimal solution obtained using a meta-heuristic is refined using a local search have also become popular.
Villaverde et al. combined scatter search with local search methods for parameter estimation in a collection of systems biology models \cite{villaverde2015biopredyn}.
Fan et al. recently showed that population based meta-heuristics along with decomposition based methods can be also used to model gene circuits from mRNA data \cite{fan2015parameter}.
Despite these successes, a major drawback of most metaheuristic approaches is the large number of function evaluations required to explore the parameter space.
Typically as models grow in size and complexity, evaluation of the objective function becomes computationally expensive.
Thus performing a large number of function evaluations is not desirable (and perhaps not feasible).

@@

Villaverde and co-workers published a set of benchmark biochemical problems to evaluate parameter estimation methods \cite{villaverde2015biopredyn}.
They ranked ordered these example problems by computational cost from most to least expensive.
We evaluated the performance of DOPS on problems from the least and most expensive categories.
The least expensive problem (henceforth referred to as problem B4) was a metabolic model of Chinese Hamster Ovary (CHO) with 35 metabolites, 32 reactions and 117 parameters \cite{villaverde2014high}.
The biochemical reactions were modeled using modular rate laws and the generalized form of Michaelis-Menten kinetics.
The expensive problem was a genome scale kinetic model of \textit{Saccharomyces~cerevisiae} with 261 reactions, 262 variables and 1759 parameters \cite {smallbone2013large}(henceforth referred to as problem B1). In both cases, synthetic time series data sets were generated using known parameter values, these data were then used by the parameter estimation methods.
For problem B1, the time series data consisted of 44 observables, and for problem B4 the data corresponded to 13 different metabolite measurement sets.
We fixed the number of function evaluations at $\mathcal{N}$ = 4000, and trained both models against the synthetic experimental data.
In both cases we found good fits  (Fig. \ref{fig-sims-b1} and Fig. \ref{fig-sims-b4}) to the problems.
We also recaptured the nominal parameter values within the specified number of function evaluations (Fig. \ref{fig-benchmark}).
Thus, DOPS estimated the parameters in benchmark biochemical models, and recovered the original parameters from synthetic data.
Next, we compared the performance of DOPS with the four other meta-heuristic approaches for a model of the human coagulation cascade.
